---
title: "Albrecht-preliminary-analysis-3"
author: Evelyn Albrecht
format: html
editor: visual
---

## Introduction

In this analysis, I am looking at the impacts that longevity, body size, generation time, home range size, and percent fruit in diet as indicators of ICUN status. To determine any relationships, I will be conducting another ordinal logistic regression. I hypothesize that species with long lifespan, long generation time, large home ranges, and high percent fruit in diet will correspond with more severe IUCN categorizations.

## Setting up

Clear environment and open relevant packages.

```{r}
rm(list=ls())  
library(tidyverse) 
library("here")
library('ordinal')
```

Read in the data.

```{r}
birds <- read.csv(here("Data","Albrecht_Frugivoria_bird_database_2023_full.csv"))
#I am using this original data because I am keeping all of the values instead of getting rid of outliers. additionally, I am keeping LC as a category so it is just the original dataset
```

### Skipping the vegan stuff (see analysis one for more in depth explanation)

#### Basically, the vegan ordination stuff for my analysis

### Trying the Geeks for Geeks method

Access the information I have been using: <https://www.geeksforgeeks.org/ordinal-logistic-regression-in-r/>

Use the MASS package(installed in console) and polr() function within that package.

```{r}
library(MASS)
```

## Create a model

Create a model with this formula.

```{r}
#example formula 
  #model <- polr(response ~ predictor1 + predictor2, data=data, Hess=TRUE)
#Hess = Hessian matrix "which is necessary for certian model diagnostics"
```

Now to try with my data...

Gotta convert IUCN category from character to factor data first.

```{r}
birds$IUCN_category <- as.factor(birds$IUCN_category)
```

Create model measuring IUCN category as a function of longevity, body size, generation time, home range size, and percent fruit in diet (and percent invertebrate in diet)

```{r}
bird_model <- polr(IUCN_category ~ longevity + body_size_mm + home_range_size + generation_time + diet_fruit_e, data = birds, Hess = TRUE)
```

"Error in polr(IUCN_category \~ longevity + body_size_mm + home_range_size + : attempt to find suitable starting values failed" - it seems to be because the insect diet variable has entries that are 0, the model will not work.

I should either remove the variable all together or omit the 0s within that variable.

Creating the model without that variable for now.

## Check the model assumptions

Due to the models nature, it does not have any assumptions it abides by that requires checking.

## Interpret the model

Gather the sats from the summary() function.

```{r}
summary(bird_model)
```

I saw that the bigger the absolute value of t, the higher impact and that values greater than 2 and less than -2 indicate significance.

Longevity, home range, and percent fruit in diet appear to not not have a significant impact on IUCN categorization.

Body size and generation time seem to have a significant impact on IUCN category.

Due to having the highest absolute value, it seems like generation time is the best indicator for IUCN category.

I also noticed the "1031 observations deleted due to missingness" and am not sure how to interpret that and whether is should be worried.

### Result Statements: I found a significant relationship between body size and IUCN category (t=2.7656) and between generation time and IUCN category (-3.5317). There was no significant relationship between Longevity, home range, and percent fruit in diet and IUCN category (t=-0.1368, t=-0.1095, t=0.8974)

## Make a new (pretty) plot

using the Geek for Geek stuff again

lets dive into it?

I am completely lost. I do not understand how to apply what is on the page to my data.
