---
title: "Albrecht-preliminary-data-analysis-1"
author: "Evelyn Albrecht"
format: html
editor: visual
---

### Introduction

In this analysis, I will be looking at the impacts of diet breadth and human footprint on IUCN category.

In order to determine significance of relationships, I will be conducting an ordinal logistic regression. I predict that larger diet breadths would result in lower IUCN categorizations and that higher human footprints would correspond with higher IUCN categories.

#### Setting up

```{r}
rm(list=ls())  
library(tidyverse) 
library("here")
library(vegan) 
library('ordinal') # package use to run an ordinal logistic regression. 
```

#### Read in the data.

```{r}
mammals <- read.csv(here("Data","Frugivoria_mammals_edited_again.csv"))
```

### Experimenting with different ways of going about ordinal stuff

#### Using Erika's tutorial qmd thing for converting data frames and ordinationx

My data frame is long, not wide. Therefore it will need to be converted into a wide one.

vertical. each species is a list. i think we want IUCN status as the first thing?

lets try and imimitate the workflow

```{r}
long_mammal_summary <- mammals %>%
  group_by(IUCN_category, diet_breadth, mean_human_fp_range_2020) %>%
  summarise(num_icun = length(IUCN_category))
```

This doesn't seem like it is working.

I don't think this is actually applicable to me cause I don't have count data, every animal is its own observation.

\*11/22/2024 note - this isn't applicable to me because I am doing ordinal logistic regression, not ordination.

## Initial plot of the data.

```{r}
ggplot(mammals, aes(x=IUCN_category, y=diet_breadth)) + 
  geom_boxplot() + 
  theme_bw()
```

```{r}
ggplot(mammals, aes(x=IUCN_category, y=mean_human_fp_range_2020)) + 
  geom_boxplot() + 
  theme_bw()
```

## Create a model

### Trying the Geeks for Geeks method

URL to what I am referencing: <https://www.geeksforgeeks.org/ordinal-logistic-regression-in-r/>

Use the MASS package(in console) and polr() function within that package.

```{r}
library(MASS)
```

Create a model with is formula.

```{r}
#example formula 
  #model <- polr(response ~ predictor1 + predictor2, data=data, Hess=TRUE)
#Hess = Hessian matrix "which is necessary for certian model diagnostics"
```

Now to try with my data

Gotta convert IUCN category from character to factor data first.

```{r}
mammals$IUCN_category <- as.factor(mammals$IUCN_category)
```

Create a model looking at IUCN category as a function of diet breadth and human footprint out of the mammal dataset.

```{r}
mammal_model <- polr(IUCN_category ~ diet_breadth + mean_human_fp_range_2020, data = mammals, Hess = TRUE)
```

YAY! the model was created!!

## Check the model assumptions

Because this is not a linear regression, there are no assumptions we need to check.

## Interpret the model

Use the summary() function to obtain statistics.

```{r}
summary(mammal_model)
```

Apparently, the bigger the t-value (positive or negative), the larger impact the numeric variable has on ordinal one. I additionally saw a general consensuses that values greater than 2 (or less than -2 if negative) indicate significance.

Working off of this knowledge, I can claim that mean human footprint has a significant affect on IUCN categories whilst diet breadth does not (although the value is relatively close to 2).

### Results statement: I found there was a significant relationship between mean human footprint and IUCN category (t=-3.017). There was no relationship between diet breadth and IUCN category (t=1.778)

## Make a new (pretty) plot

I am still following the Geeks for Geeks site instructions but I am very very lost.

lets dive into it?

Here is the example the site provides. I am lost...

```{r}
# Create a sequence of scores for prediction
new_data <- data.frame(Score = seq(min(data$Score), max(data$Score), length.out = 100))
 
# Predict probabilities for each category
predicted_probs <- as.data.frame(predict(model, newdata = new_data, type = "probs"))
 
# Rename the columns for clarity
colnames(predicted_probs) <- c("1", "2", "3", "4", "5")
 
# Create a visualization
ggplot(predicted_probs, aes(x = new_data$Score)) +
  geom_line(aes(y = `1`, color = "Category 1"), size = 1) +
  geom_line(aes(y = `2`, color = "Category 2"), size = 1) +
  geom_line(aes(y = `3`, color = "Category 3"), size = 1) +
  geom_line(aes(y = `4`, color = "Category 4"), size = 1) +
  geom_line(aes(y = `5`, color = "Category 5"), size = 1) +
  labs(title = "Ordinal Logistic Regression",
       x = "Score",
       y = "Predicted Probability") +
  scale_color_manual(values = c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd"),
                     name = "Category") +
  theme_minimal()
```
