---
title: "Albrecht-preliminary-analysis-2"
author: "Evelyn Albrecht"
format: html
editor: visual
---

#### Introduction

In this analysis, I am looking at the relationship between diet breadth and body size of 118 parrot species in the neotropics.

Given that both variables are ratio data, a linear regression is required to determine if there is a correlation between these two factors.

I predict that there will be a positive relationship in diet breadth and body size in neotropical parrots.

#### Setting up

Clear R's environment and open relevant packages.

```{r}
rm(list = ls())
library(tidyverse)
library("here")
library(ggfortify)
```

Read in the data and assign it the name "parrots"

```{r}
parrots <- read.csv(here("Data", "frugivoria_parrots.csv"))
```

Create an initial plot of the data comparing body size and diet breadth.

```{r}
ggplot(parrots, aes(x=diet_breadth, y=body_size_mm)) + 
  geom_point() + 
  xlab("Diet breadth") + 
  ylab("Body size (mm)") + 
  theme_bw()
```

It does not look like there is much of a correlation between diet breadth and body size.

Lets try to guesstimate a theorhetical slope a line of best fit anyway.

Diet breadth seems to range from 0 to about 1.4

Body size ranges from about 20(?) to 950 mm

lets use these ranges to calculate a *very* rough estimate of a slope.

y2 - y1 / x2 - x1

(950-20)/(1.4-0)

Our estimated slope is 930/1.4 = 664.286. This means with each increase of 1 diet breadth metric, we anticipate

Its very difficult to pinpoint a y-intercept with something so disorganized. I think it looks to be close to zero.

## Create model

create a model that measures body size as a function of diet breadth.

```{r}
parrot_model <- lm(body_size_mm ~ diet_breadth, data = parrots)
```

## Check model assumptions

Check the assumptions of our model using autoplot.

```{r}
autoplot(parrot_model)
```

That does not look good. The Normal Q-Q plot is very off, very abnormal. Our data does not meet the assumptions our general linear regression model dictates.

I will have to rethink about how we approach this analysis. I am going to move forwards with the usual workflow for now.

Time to interpret the model.

## Interpret the model

Run an anova and summary to gather statistics.

anova() first:

```{r}
anova(parrot_model)
```

Yikes. Continue onto the summary().

Our F=value is large and p-value was large (\>0.05) and therefore we are unable to reject our null hypothesis that there is a correlation between diet breadth and body size.

```{r}
summary(parrot_model)
```

From this test we obtain really small F-value(0.03552) and large p-value (0.8488) indicate that there was no significant relationship between the two variables. The adjusted R-squared value should indicate how much variation in body size is associated with diet breadth. My Adjusted R-squared value is very small and negative. I have no clue what exactly this indicates. Google is telling me there is probably an error in my model so I wonder if it has to do with the model not meeting at assumptions.

### Results Statement: I found no significant relationship between diet breadth and body size (df =1, F= 0.0365, p = 0.8488)

## Make a new (pretty) plot

```{r}
ggplot(parrots, aes(x=diet_breadth, y=body_size_mm)) + 
  geom_point(color = "darkred") + 
  geom_smooth(method = "lm") + 
  xlab("Diet breadth") + 
  ylab("Body size (mm)") + 
  theme_bw()
```
